---
title: "Introduction to corrmorant"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to corrmorant}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

```{r setup, include = FALSE}
library(corrmorant)
library(dplyr)
```


## Description
`corrmorant` extends `ggplot2` by an automated framework for plots of correlation matrices that can be easily modified via regular `ggplot2` syntax. In addition, it provides a large set of visualization tools for exploratory data analysis based on correlation matrices.

A big thank you goes out to the developers of the software without which this package would not be possible, especially [Hadley Wickham](https://github.com/hadley/) and the rest of the [ggplot2](https://github.com/tidyverse/ggplot2) development team for their incredible work.

Here`s a list of everyone else directly or indirectly implied in this package - thanks a lot for your invaluable contribution to the R community:

``` {r}
# Contributers to the packages used in corrmorant
thankr::shoulders(package = "corrmorant") %>% 
  # Remove number of packages 
  select(-no_packages) %>% 
  # Remove myself
  dplyr::filter(!grepl("Roman M. Link", maintainer)) %>% 
  # Remove email addresses 
  mutate(maintainer = stringr::word(maintainer, end = -2)) %>% 
  # Return as kable
  knitr::kable()
```


## Motivation
The inspection of correlation matrices is a central part of exploratory data analysis. As collinearity between predictors has the potential to seriously compromise the quality of regression results - as well as hinder a straightforward interpretation of regression coefficients - the inspection of the correlations among potential predictor variables as well as the desired response should precede fitting any regression model with more than one predictor. 

If you are using Bayesian methods, you will also often be faced with the necessity to inspect the correlation between parameters estimated from MCMC samples, which means you have to effectively visualize correlations in often huge datasets.

In addition, with structural equation models (and their cousins, confirmatory factor analysis and path analysis), there is an entire class of models that traditionally is estimated directly from the covariance matrix of a multivariate dataset. 

Long story short - there are plenty of reasons why one might want to visually inspect the interrelationships between all variables in a dataset simultaneously. However, while R provides plenty of functions and packages to display correlations, so far (at least to my knowledge) all of them are only able to display certain types of correlation plots, and none of them truly leverages the full flexibility of `ggplot2`. 

`corrmorant` is meant to give you this flexibility. It is designed to blend in seamlessly in a data analysis framework based on the `tidyverse`. It offers a broad variety of functions that allow you to create a large number of different correlation plots, and can be extended easily with new geoms, stats and themes, just as `ggplot2` itself. 

In case you wonder about the name: One day when I was riding my bike to work and thinking about stupid puns related to **corr**elation **m**atrices, a shiny black bird appeared...


## Installation
The package can be installed from Github using `devtools::install_github()`:

``` {r, eval = FALSE}
# install devtools package if necessary
install.packages("devtools")
# install corrmorant from the github repository
devtools::install_github("r-link/corrmorant")
```

Afterwards, the package can be loaded regularly via `library()`:
``` {r}
library(corrmorant)
```


## Simple plots with corrmorant()
The `corrmorant()` function is a simple wrapper function around the more complex `gcorrm()` function that can be used to create first, simple plots of correlation matrices. Currently, 
three different styles are available, "light", "dark" and "blue_red":

``` {r}
# correlation plot of the iris data using style = 'light'
corrmorant(iris, style = "light")
# the "dark" style has a dark background in the diagonal facets
corrmorant(iris, style = "dark")
# in the "blue_red" style, colors and correlation labels are colored by the strength of
# correlation
corrmorant(iris, style = "blue_red")
```

## Basic usage of ggcorrm()
While `corrmorant()` may be useful for many basic data inspection tasks, its functionality is rather limited. If you want to take control of the elements in a plot, you can use `ggcorrm()` together with the corrmorant selector functions `lotri()`, `utri()` and `dia()` that direct ggplot layers to the lower or upper triangle or the plot diagonal of a `ggcorrm()` plot, respectively. In addition, corrmorant offers a series of utility functions for readymade data summaries on the plot diagonal (`dia_names()`, `dia_density()`, `dia_histogram` and `dia_freqpoly()`) as well as a couple of new stats, e.g. `stat_corrtext()` for displaying correlation strength in correlation plot facets. The new `corrmorant` stats can generally be called in a simplified form by prefixing their name with `lotri_` or `utri_`, e.g. `lotri_corrtext()` and `utri_corrtext()`

For example, `corrmorant(iris, style = "light")` can be recreated by the following code:

```{r}
p1 <- ggcorrm(iris) +
  lotri(geom_point(alpha = 0.5)) +
  utri_corrtext() +
  dia_names(y_pos = 0.1, size = 3) +
  dia_density(lower = 0.4, fill = "grey80", color = 1)
p1
```

Additional layers can than be easily added using`corrmorant` selector functions:

```{r}
p1 + lotri(geom_smooth(method = "lm"))
```

## Working principle of ggcorrm()
### Data preparation
Under the hood, ggcorrm arranges the facets of the correlation matrix using `ggplot2::facet_grid()`. To be able to arrange a dataset across the facets, as a first step, `ggcorrm()` converts the data in a long-table format suitable for plotting. 

This is achieved by calling a `tidy_corrm` method, which automatically rearranges the columns in a suitable way and returns them in a tidy long form in a `tidy_corrm` object (a special type of tibble):

``` {r}
tidy_corrm(iris)
```

As seen in the given example, there is a number of standard column names that describe the content of all numeric variables in the facets (`var_x:corr_group`), while all categorical variables are carried over with the dataset in their original form and can be used for grouping (in this case only `Species`). Additional columns can be created from these columns using the `mutates` argument (see below).


The content of the columns in a `tidy_corrm` object is as follows:

------------- -----------------------------------------------------------------
`var_x`       Name of the variable on the x-axis in the order of
              appearance in the raw data (ordered factor).

`var_y`       Name of the variable on the y-axis in the order of appearance
              in the raw data (ordered factor).

`x`           Data of the variable on the x axis (numeric).

`y`           Data of the variable on the y axis (numeric).

`type`        Type of panel (character, `"upper"`, `"lower"` or `"diag"`).

`.corr`       Correlation between x and y for the respective panel/group,
              `stats::cor()` using the method specified by
              `corr_method` and optionally within the groups specified with 
              `corr_group` (numeric).

`corr_group`  grouping variable for `.corr` (1 for all observations if
              no groups are specified).

Additional    All other columns specified in the dataset and/or 
columns       created via `mutates`.
------------- -----------------------------------------------------------------

It is also possible to call `ggcorrm()` directly on a `tidy_corrm` object. In this case, it will be directly plotted without reshaping of the data, and arguments that `ggcorrm()` passes to  `tidy_corrm()` are ignored. 

`tidy_corrm()` provides a set of options;
* labels
* correlation
* grouping
* mutates

