---
title: "Introduction to corrmorant"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to corrmorant}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

```{r setup, include = FALSE}
library(corrmorant)
library(dplyr)
```


## Description
`corrmorant` extends `ggplot2` by an automated framework for plots of correlation matrices that can be easily modified via regular `ggplot2` syntax. In addition, it provides a large set of visualization tools for exploratory data analysis based on correlation matrices.

## Motivation
The inspection of correlation matrices is a central part of exploratory data analysis. As collinearity between predictors has the potential to seriously compromise the quality of regression results - as well as hinder a straightforward interpretation of regression coefficients - the inspection of the correlations among potential predictor variables as well as the desired response should precede fitting any regression model with more than one predictor. 

If you are using Bayesian methods, you will also often be faced with the necessity to inspect the correlation between parameters estimated from MCMC samples, which means you have to effectively visualize correlations in often huge datasets.

In addition, with structural equation models (and their cousins, confirmatory factor analysis and path analysis), there is an entire class of models that traditionally is estimated directly from the covariance matrix of a multivariate dataset. 

Long story short - there are plenty of reasons why one might want to visually inspect the interrelationships between all variables in a dataset simultaneously. However, while R provides plenty of functions and packages to display correlations, so far (at least to my knowledge) all of them are only able to display certain types of correlation plots, and none of them truly leverages the full flexibility of `ggplot2`. 

`corrmorant` is meant to give you this flexibility. It is designed to blend in seamlessly in a data analysis framework based on the `tidyverse`. It offers a broad variety of functions that allow you to create a large number of different correlation plots, and can be extended easily with new geoms, stats and themes, just as `ggplot2` itself. 

In case you wonder about the name: One day when I was riding my bike to work and thinking about stupid puns related to **corr**elation **m**atrices, a shiny black bird appeared...


## Installation
The package can be installed from Github using `devtools::install_github()`:

``` {r, eval = FALSE}
# install devtools package if necessary
install.packages("devtools")
# install corrmorant from the github repository
devtools::install_github("r-link/corrmorant")
```

Afterwards, the package can be loaded regularly via `library()`:
``` {r}
library(corrmorant)
```


## Simple plots with corrmorant()
The `corrmorant()` function is a simple wrapper function around the more complex `gcorrm()` function that can be used to create first, simple plots of correlation matrices. Currently, three different styles are available, "light", "dark" and "blue_red":

``` {r}
# get subset of drosera dataset for Drosera capensis
capensis <- filter(drosera, species == "capensis")
# correlation plot of the drosera data using style = 'light'
corrmorant(capensis, style = "light")
# the "dark" style has a dark background in the diagonal facets
corrmorant(capensis, style = "dark")
# in the "blue_red" style, colors and correlation labels are colored by the strength of
# correlation
corrmorant(capensis, style = "blue_red")
```

The`corrmorant()` interface is kept simple on purpose as this function is only intended for fast visual checks of datasets. If you want to modify single elements of the plot, it is always preferable to use `ggcorrm()` instead.


## Basic usage of ggcorrm()
The `ggcorrm()` function allows to create correlation matrices with a syntax that gently extends regular `ggplot()` syntax. If calling `ggcorrm()` on a data.frame, it automatically rearranges it into the form needed for plotting, treating all numeric variables as columns/rows of the correlation matrix, and retains all other data as additional columns of the plotting dataset that can be used for grouping. 

### Adding regular ggplot2 layers
You can add new layers and other `ggplot2` plot elements (e.g. scales, themes etc.) to a `ggcorrm()` object as if it was a regular `ggplot()` call. However, adding layers to all panels of a correlation matrix is rarely useful:

```{r}
ggcorrm(capensis) + geom_point(alpha = 0.5)
```

Usually, you will want to display different graphical elements in the plot diagonal, lower and upper triangular panels of the correlation matrix, respectively. This can be achieved with the three corrmorant selector functions `lotri()`, `utri()` and `dia()`, which can be used to modify ggplot layers (i.e. the output of calls to geoms, stats or the `layer()` function) and direct them only to the lower or upper triangle or the plot diagonal of a `ggcorrm()` plot, respectively. To use them, you simply have to wrap the ggplot layer into one of the selector functions:

```{r}
ggcorrm(capensis) + 
  lotri(geom_point(alpha = 0.5))
```

### New corrmorant stats
In addition to the default geoms and stats available in `ggplot2`, corrmorant introduces a series of new stats that are specifically geared towards the creation of correlation matrices. These can be grouped into a) stats for data summaries and information on the plot diagonal and b) stats for data display in the off-diagonal panels. To simplify their usage, all corrmorant stats can be called directly by prefixing their name with `dia_`, `lotri_` or `utri_` instead of calling them via the corresponding selector functions. The prefixed versions differ from calls to the underlying stats by having a set of reasonable standard values and are preferable in most cases.

**corrmorant stats for the plot diagonal**

---------------------- --------------------- ------------------------------------
Stat                   Prefixed version      Description
---------------------- --------------------- ------------------------------------
`stat_dia_names()`     `dia_names()`         display text labels for variable 
                                             names on the plot diagonal

`stat_dia_density()`   `dia_density()`       display density plots in diagonal
                                             panels

`stat_dia_histogram()` `dia_histogram()`     display histograms in diagonal
                                             panels

`stat_dia_freqpoly()`  `dia_freqpoly()`      display frequency polygons in 
                                             diagonal panels
---------------------- --------------------- ------------------------------------


**corrmorant stats for off-diagonal panels**

---------------------- --------------------- ------------------------------------
Stat                   Prefixed versions     Description
---------------------- --------------------- ------------------------------------
`stat_corrtext()`      `lotri_corrtext()`    place text labels indicating  
                       `utri_corrtext()`     correlation strength 
                       
`stat_funtext()`       `lotri_funtext()`     place text labels based on                                   
                       `utri_funtext()`      user-defined functions  

`stat_heatmap()`       `lotri_heatmap()`     add correlation heatmap 
                       `utri_heatmap()`      
                       
`stat_heatpoint()`     `lotri_heatpoint()`   add symbols whose size and color  
                       `utri_heatpoint()`    indicates correlation strength

`stat_heatcircle()`    `lotri_heatcircle()`  add circles whose area scales with
                       `utri_heatcircle()`   correlation strength
---------------------- --------------------- ------------------------------------

Together with regular `ggplot2` geoms and stats, these new functions can be used as building blocks for more complex correlation matrix plots.

For example, `corrmorant(capensis, style = "light")` can be recreated by the following code:

```{r}
ggcorrm(capensis) +
  lotri(geom_point(alpha = 0.5)) +
  utri_corrtext() +
  dia_names(y_pos = 0.1, size = 3) +
  dia_density(lower = 0.4, fill = "grey80", color = 1)
```




## Working principle of ggcorrm()
### Data preparation
Under the hood, ggcorrm arranges the facets of the correlation matrix using `ggplot2::facet_grid()`. To be able to arrange a dataset across the facets, as a first step, `ggcorrm()` converts the data in a long-table format suitable for plotting. 

This is achieved by calling a `tidy_corrm` method, which automatically rearranges the columns in a suitable way and returns them in a tidy long form in a `tidy_corrm` object (a special type of tibble):

``` {r}
tidy_corrm(drosera)
```

As seen in the given example, there is a number of standard column names that describe the content of all numeric variables in the facets (`var_x:corr_group`), while all categorical variables are carried over with the dataset in their original form and can be used for grouping (in this case only `Species`). Additional columns can be created from these columns using the `mutates` argument (see below).


The content of the columns in a `tidy_corrm` object is as follows:

------------- -----------------------------------------------------------------
Column        Content
------------- -----------------------------------------------------------------
`var_x`       Name of the variable on the x-axis in the order of
              appearance in the raw data (ordered factor).

`var_y`       Name of the variable on the y-axis in the order of appearance
              in the raw data (ordered factor).

`x`           Data of the variable on the x axis (numeric).

`y`           Data of the variable on the y axis (numeric).

`type`        Type of panel (character, `"upper"`, `"lower"` or `"diag"`).

`.corr`       Correlation between x and y for the respective panel/group,
              `stats::cor()` using the method specified by
              `corr_method` and optionally within the groups specified with 
              `corr_group` (numeric).

`corr_group`  grouping variable for `.corr` (1 for all observations if
              no groups are specified).

Additional    All other columns specified in the dataset and/or 
columns       created via `mutates`.
------------- -----------------------------------------------------------------

It is also possible to call `ggcorrm()` directly on a `tidy_corrm` object. In this case, it will be directly plotted without reshaping of the data, and arguments that `ggcorrm()` passes to  `tidy_corrm()` are ignored. 

`tidy_corrm()` provides a set of options;
* labels
* correlation
* grouping
* mutates

