---
title: "Introduction to corrmorant"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to corrmorant}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

```{r setup, include = FALSE}
library(corrmorant)
library(dplyr)
```


## Description
`corrmorant` extends `ggplot2` by an automated framework for plots of correlation matrices that can be easily modified via regular `ggplot2` syntax. In addition, it provides a large set of visualization tools for exploratory data analysis based on correlation matrices.

A big thank you goes out to the developers of the software without which this package would not be possible, especially [Hadley Wickham](https://github.com/hadley/) and the rest of the [ggplot2](https://github.com/tidyverse/ggplot2) development team for their incredible work.

Here`s a list of everyone else directly or indirectly implied in this package - thanks a lot for your invaluable contribution to the R community:

``` {r}
# Contributers to the packages used in corrmorant
thankr::shoulders(package = "corrmorant") %>% 
  # Remove number of packages 
  select(-no_packages) %>% 
  # Remove myself
  dplyr::filter(!grepl("Roman M. Link", maintainer)) %>% 
  # Remove email addresses 
  mutate(maintainer = stringr::word(maintainer, end = -2)) %>% 
  # Return as kable
  knitr::kable()
```


## Motivation
The inspection of correlation matrices is a central part of exploratory data analysis. As collinearity between predictors has the potential to seriously compromise the quality of regression results - as well as hinder a straightforward interpretation of regression coefficients - the inspection of the correlations among potential predictor variables as well as the desired response should precede fitting any regression model with more than one predictor. 

If you are using Bayesian methods, you will also often be faced with the necessity to inspect the correlation between parameters estimated from MCMC samples, which means you have to effectively visualize correlations in often huge datasets.

In addition, with structural equation models (and their cousins, confirmatory factor analysis and path analysis), there is an entire class of models that traditionally is estimated directly from the covariance matrix of a multivariate dataset. 

Long story short - there are plenty of reasons why one might want to visually inspect the interrelationships between all variables in a dataset simultaneously. However, while R provides plenty of functions and packages to display correlations, so far (at least to my knowledge) all of them are only able to display certain types of correlation plots, and none of them truly leverages the full flexibility of `ggplot2`. 

`corrmorant` is meant to give you this flexibility. It is designed to blend in seamlessly in a data analysis framework based on the `tidyverse`. It offers a broad variety of functions that allow you to create a large number of different correlation plots, and can be extended easily with new geoms, stats and themes, just as `ggplot2` itself. 

In case you wonder about the name: One day when I was riding my bike to work and thinking about stupid puns related to **corr**elation **m**atrices, a shiny black bird appeared...


## Installation
The package can be installed from Github using `devtools::install_github()`:

``` {r, eval = FALSE}
# install devtools package if necessary
install.packages("devtools")
# install corrmorant from the github repository
devtools::install_github("r-link/corrmorant")
```

Afterwards, the package can be loaded regularly via `library()`:
``` {r}
library(corrmorant)
```


## Simple plots with corrmorant()
The `corrmorant()` function is a simple wrapper function around the more complex `gcorrm()` function that can be used to create first, simple plots of correlation matrices. Currently, 
three different styles are available, "light", "dark" and "blue_red":

``` {r}
# correlation plot of the iris data using style = 'light'
corrmorant(iris, style = "light")
# the "dark" style has a dark background in the diagonal facets
corrmorant(iris, style = "dark")
# in the "blue_red" style, colors and correlation labels are colored by the strength of
# correlation
corrmorant(iris, style = "blue_red")
```

## Basic usage of ggcorrm()
While `corrmorant()` may be useful for many basic data inspection tasks, its functionality is rather limited. If you want to take control of the elements in a plot, you can use `ggcorrm()` together with the corrmorant selector functions `lotri()`, `utri()` and `dia()` that direct ggplot layers to the lower or upper triangle or the plot diagonal of a `ggcorrm()` plot, respectively. In addition, corrmorant offers a series of utility functions for readymade data summaries on the plot diagonal (`dia_names()`, `dia_density()`, `dia_histogram` and `dia_freqpoly()`) as well as a couple of new stats, e.g. `stat_corrtext()` for displaying correlation strength in correlation plot facets. The new `corrmorant` stats can generally be called in a simplified form by prefixing their name with `lotri_` or `utri_`, e.g. `lotri_corrtext()` and `utri_corrtext()`

For example, `corrmorant(iris, style = "light")` can be recreated by the following code:

```{r}
p1 <- ggcorrm(iris) +
  lotri(geom_point(alpha = 0.5)) +
  utri_corrtext() +
  dia_names(y_pos = 0.1, size = 3) +
  dia_density(lower = 0.4, fill = "grey80", color = 1)
p1
```

If you want, you can add additional ggplot2 layers, using the appropriate corrmorant selectors to restrict them to the appropriate set of facets. For example, you might want to display a linear trend in the facets on the lower triangle:

```{r}
p1 + lotri(geom_smooth(method = "lm"))
```

The horrific fit of these linear regressions results from the fact that the iris dataset contains data from three different species. You might want to include this information into your plot by plotting the three species in different colours, which can easily be achieved by setting plot level aesthetics using the `mapping` argument of `ggcorrm()`:

```{r}
ggcorrm(iris, mapping = aes(col = Species, fill = Species)) +
  lotri(geom_smooth(method = "lm")) +
  lotri(geom_point(alpha = 0.5)) +
  utri_corrtext(nrow = 2, squeeze = 0.6) +
  dia_names(y_pos = 0.1, size = 3) +
  dia_density(lower = 0.4, color = 1)
```

## Coloring data by correlation strength
To color a dataset by correlation strength, you can use the `.corr` column internally created in `tidy_corrm()` together with `scale_colour_corr()` or `scale_fill_corr()`:

```{r}
# simulate some data from a multivariate normal distribution
# (mtcars is only used as a shortcut to a positive definite covariance matrix)
data1 <- MASS::mvrnorm(100, colMeans(mtcars),  cov(mtcars))[, 1:8]
colnames(data1) <- paste("Var.", 1:ncol(data1))

# create plot
ggcorrm(data1, 
        mapping = aes(col = .corr, fill = .corr),
        bg_dia = "grey20") +
  lotri(geom_smooth(method = "lm", size = .3)) +
  lotri(geom_point(alpha = 0.5)) +
  utri_corrtext(nrow = 2, squeeze = 0.6) +
  dia_names(y_pos = 0.1, size = 3, color = "white") +
  dia_histogram(lower = 0.4, color = "grey80", fill = "grey60", size = .3) +
  scale_color_corr(aesthetics = c("fill", "color"))
```

## Correlation heatmaps and more 
`corrmorant` offers a series of new geoms and stats that are designed to improve the display of correlation strength.

For example, there is a set of stats for correlation heatmaps and the likes, which can be very useful when inspecting datasets with large numbers of variables:

```{r}
select(mtcars, mpg, disp:qsec) %>% 
ggcorrm() +
  utri_heatmap(alpha = 0.5) +
  lotri_heatcircle(alpha = 0.5, col = 1) +
  utri_corrtext() +
  dia_names(y_pos = 0.1, size = 3) +
  dia_density(lower = 0.4, fill = "lightgrey", color = 1) +
  scale_fill_corr() 
```


With `lotri_/utri_heatpoint()`, you can scale the diameter of any character you want by correlation strength.
And yes, your right, this means you can display correlation strength with purple skulls and angry cat faces:

```{r, warning = FALSE}
airquality %>% 
ggcorrm(aes(col = .corr)) +
  lotri_heatpoint(pch = "\U1F63E") +
  utri_heatpoint(pch = "\U2620", col = "#660066") +
  dia_names(y_pos = 0.1, size = 3) +
  dia_density(lower = 0.4, fill = "#89DFA3", color = 1) +
  scale_size(range = c(1, 15)) + 
  scale_color_corr(option = "C")
```


## Print output of arbitrary functions of the data
If you want to place text labels for the output of arbitrary functions in your plots, you can use `lotri/utri_funtext()`, which uses the same placement rules as `stat_corrtext()`.

For instance, you could add lables for the slope of the linear models in the lower triangle (note how y and x are swapped to account for the reverse axes in the upper triangle).

```{r}
# function to compute linear model slopes
lmslope <- function(x, y)  round(coef(lm(x ~ y))[2], 2)

# add slopes using a function
ggcorrm(iris, mapping = aes(col = Species, fill = Species)) +
  lotri(geom_point(alpha = 0.4)) +
  lotri(geom_smooth(alpha = 0.4, method = "lm")) +
  utri_funtext(fun = lmslope, squeeze = .6) +
  dia_density(lower = .4, col = 1, alpha = 0.4) +
  dia_names(y_pos = .1)
```

Besides using a two-parameter formula, `stat_funtext()`, the stat underlying `lotri/utri_funtext()`, also accepts rlang style lambda expressions describing functions and quosures generated with `quo()` (referring to the variable names `x` and `y`).

Here are two ways to copy the output of `stat_corrtext()` - using a lambda expression (a formula containing `.x` and `.y`) in the upper and a quosure in the lower triangle:

```{r}
ggcorrm(iris, aes(col = .corr), rescale = "as_is") +
  utri_funtext(fun = ~ round(cor(.x, .y), 2)) +
  lotri_funtext(fun = quo(round(cor(x, y), 2))) +
  dia_names(y_pos = .5) +
  scale_color_corr()
```

